The Array of Long Baseline Antennas for Taking Radio Observations from the Sub-Antarctic

### 2020-10-29 albatros_paper.tex
added scinet acknowledgement
HCC updated some responses

Very few experiments have surveyed the radio sky at $\lesssim 30$~MHz,
and here, we highlight only a subset.  At the very lowest frequencies,
the state of the art among ground-based measurements dates from the
1950s, when Reber and Ellis caught brief glimpses of the $2.1$~MHz sky
at $\sim 5$\degree\ resolution, using an array of
192~dipoles~\citep{1956JGR....61....1R}.  A few space-based missions
have also performed measurements at similarly low frequency ranges;
for example, the Radio Astronomy Explorer-2 operated at
\SIrange{0.025}{13}{\MHz} with $\sim 10$\degree\ resolution at
\SI{4.7}{\MHz}~\citep{1975A&A....40..365A}.  More recent ground-based
experiments include the Dominion Radio Astrophysical Observatory
surveys of the northern sky at 10~MHz and 22~MHz, with
$\sim2$\degree\ and 1.1--1.7\degree\ resolution,
respectively~\citep{1976MNRAS.177..601C, 1999A&AS..137....7R}.
Low-frequency imaging experiments that are currently operational or
under construction include the Ukranian T-shaped Radio
telescope~\citep{2016JAI.....541010Z}, with 0.5\degree\ resolution at
the center of its 8--32~MHz operating range, and the New extension in
Nan\c{c}ay upgrading LOFAR~\citep{7136773}, which operates at
10--85~MHz with 1.5\degree\ resolution in standalone mode, and
improved resolution at $>30$~MHz when combined with LOFAR.  Existing
measurements with finer resolution are mainly at higher frequencies;
for example, the Owens Valley Long Wavelength Array mapped the sky
with \SI{15}{\arcminute} resolution between 36.528~MHz and
73.152~MHz~\citep{2018AJ....156...32E}.  Although this experimental
list is not comprehensive, it does illustrate the dearth of
information about the $\lesssim 30$~MHz sky and the lack of high
resolution measurements at the lowest frequencies, particularly in the
southern hemisphere.  The first step toward laying the groundwork for
possible measurements of the dark ages is obtaining an improved map of
Galactic foregrounds at $\lesssim 30$~MHz.  In addition to serving as
a stepping stone for future cosmological constraints from the dark
ages, maps at these frequencies can also shed new light on Galactic
astrophysics.  Above $\sim30$~MHz, synchrotron emission from the
Galaxy follows a power law with a frequency dependence of $\sim
\nu^{-0.7}$.  At lower frequencies, synchrotron self-absorption
becomes non-negligible, and the spectrum transitions to $\nu^{+5/2}$
dependence.  Since self-absorption causes the optical depth along the
line of sight to depend on frequency, observations at $\lesssim
30$~MHz have the potential to probe the three-dimensional cosmic ray
structure of the Galaxy~\citep{2002ApJ...575..217P}.  Low-frequency
observations can also be used to image radio recombination lines
(RRLs), which provide a means for studying cool, low-density regions
of the interstellar medium (ISM).  These absorption lines arise from
Rydberg atoms and are highly sensitive to the surrounding environment.
The RRL spectrum and absorption profiles can therefore constrain the
detailed properties of the ISM regions in which the Rydberg atoms are
formed~\citep{2009NewAR..53..259G, 2007MNRAS.374..852S}.  Finally,
observations at $\lesssim30$~MHz may provide new views of Earth's
ionosphere, which absorbs and refracts at radio frequencies and
becomes completely opaque below the plasma cutoff frequency.  The
cutoff frequency and the levels of absorption and refraction are
time-varying and spatially dependent.  Experiments that image the
radio sky at multiple frequencies bracketing the plasma cutoff can
therefore simultaneously image the ionosphere and provide spectral
riometry data for space weather studies~\citep{2014GeoRL..41.5370K}.

@@

The back-end readout electronics are housed in a Faraday cage located
50~m away from the antennas. The analog signal chain consists of a
pair of high- and low-pass filters (Mini-Circuits ZFHP-1R2+ and
SLP-150+) that together band-limit the signal to 1.2--140~MHz, and the
filters are followed by a Mini-Circuits ZFL-500+ amplifier.  In
contrast to the two-element pathfinder, the low-pass cutoff is
increased from 81~MHz to 140~MHz to capture downlink signals at
137--138~MHz from the ORBCOMM satellite constellation.  For the final
\albatros\ array, the ORBCOMM signals provide a convenient means for
synchronization across the antenna stations, serving as a backup to
the GPS timing discussed below.  Preliminary lab tests suggest that on
time scales of $\sim30$~s, relative timing between different antenna
stations can be measured to an accuracy of better than a few tenths of
a nanosecond using a single satellite.  The corresponding phase error
at 10~MHz is $\lesssim1^\circ$.  With publicly available orbits and
multiple satellites typically within the field of view, we expect that
ORBCOMM baseband data saved at the same time as the astronomical data
can provide offline synchronization of the \albatros\ stations.
Because the ORBCOMM and science data are taken at the same time by the
same system, improvements to the ORBCOMM calibration can be
immediately carried over into improving the science data calibration,
even if said improvements take years to realize.

As with the two-element pathfinder, a SNAP board digitizes each of the
two RF signals at 250~Msamp/s. In the autonomous station
configuration, the SNAP board ADCs are locked to a 10-MHz reference
produced by a Trimble Thunderbolt~E GPS-disciplined clock module.  The
SNAP board FPGA computes two data products: 1) channelized baseband
data for each polarization over tunable frequency windows within the
0--125~MHz operating range, with the options of 1-, 2-, or 4-bit
compression, and 2) auto- and cross-spectra from the two polarizations
over the full 0--125~MHz span, accumulated over few-second intervals.
The reduction in bit depth happens only \textit{after} the SNAP board
has channelized the baseband, and so cross-channel leakage (due to
\textit{e.g.} RFI and slopes in RF power as seen by the ADCs) is
unaffected by the low bit depth.  
An RPi~3B+ controls the SNAP board and and receives the auto- and
cross-spectra via GPIO connections, and the spectra are saved to an
on-board SD card.  The baseband data are passed from the SNAP board to
the RPi via ethernet and written to external hard drives.  The
introduction of gigabit ethernet with the RPi~3B+ model has enabled
the high data throughput associated with writing baseband.  As a
benchmark, 1-bit baseband recording of two polarizations over 10~MHz
of bandwidth yields an approximate data rate of 5~MB/s, or 0.4~TB/day.

>>>>

Very few experiments have surveyed the radio sky at $\lesssim 30$~MHz,
and here, we highlight only a subset.  At the very lowest frequencies,
the state of the art among ground-based measurements dates from the
1950s, when Reber and Ellis caught brief glimpses of the $2.1$~MHz sky
at $\sim 5$\degree\ resolution, using an array of
192~dipoles~\citep{1956JGR....61....1R}.  A few space-based missions
have also performed measurements at similarly low frequency ranges;
for example, the Radio Astronomy Explorer-2 operated at
\SIrange{0.025}{13}{\MHz} with $\sim 10$\degree\ resolution at
\SI{4.7}{\MHz}~\citep{1975A&A....40..365A}.  Subsequent ground-based
experiments include the Dominion Radio Astrophysical Observatory
(DRAO) surveys of the northern sky at 10~MHz and 22~MHz, with
$\sim2$\degree\ and 1.1--1.7\degree\ resolution,
respectively~\citep{1976MNRAS.177..601C, 1999A&AS..137....7R}.
Low-frequency imaging experiments that are currently operational or
under construction include the Ukranian T-shaped Radio
telescope~\citep{2016JAI.....541010Z}, with 0.5\degree\ resolution at
the center of its 8--32~MHz operating range, and the New extension in
Nan\c{c}ay upgrading LOFAR~\citep{7136773}, which operates at
10--85~MHz with 1.5\degree\ resolution in standalone mode, and
improved resolution at $>30$~MHz when combined with LOFAR.  Existing
measurements with finer resolution are mainly at higher frequencies;
for example, the Owens Valley Long Wavelength Array mapped the sky
with \SI{15}{\arcminute} resolution between 36.528~MHz and
73.152~MHz~\citep{2018AJ....156...32E}.  The dearth of information
about the $\lesssim 30$~MHz sky, particularly in the southern
hemisphere, is illustrated by the Global Sky
Model~\citep{2017MNRAS.464.3486Z}, which incorporates the DRAO 10~MHz
and 22~MHz data as the only two inputs at these low frequencies with
sufficiently meaningful sky coverage and resolution.
% Although this experimental list is not comprehensive, it does
% illustrate the dearth of information about the $\lesssim 30$~MHz sky
% and the lack of high resolution measurements at the lowest
% frequencies, particularly in the southern hemisphere.

The first step toward laying the groundwork for
possible measurements of the dark ages is obtaining an improved map of
Galactic foregrounds at $\lesssim 30$~MHz.  In addition to serving as
a stepping stone for future cosmological constraints from the dark
ages, maps at these frequencies can also shed new light on Galactic
astrophysics.  Above $\sim30$~MHz, synchrotron emission from the
Galaxy follows a power law with a frequency dependence of $\sim
\nu^{-0.7}$.  At lower frequencies, synchrotron self-absorption
becomes non-negligible, and the spectrum transitions to $\nu^{+5/2}$
dependence.  Since self-absorption causes the optical depth along the
line of sight to depend on frequency, observations at $\lesssim
30$~MHz have the potential to probe the three-dimensional cosmic ray
structure of the Galaxy~\citep{2002ApJ...575..217P}.  Low-frequency
observations can also be used to image radio recombination lines
(RRLs), which provide a means for studying cool, low-density regions
of the interstellar medium (ISM).  These absorption lines arise from
Rydberg atoms and are highly sensitive to the surrounding environment.
The RRL spectrum and absorption profiles can therefore constrain the
detailed properties of the ISM regions in which the Rydberg atoms are
formed~\citep{2009NewAR..53..259G, 2007MNRAS.374..852S}.  Finally,
observations at $\lesssim30$~MHz may provide new views of Earth's
ionosphere, which absorbs and refracts at radio frequencies and
becomes completely opaque below the plasma cutoff frequency.  The
cutoff frequency and the levels of absorption and refraction are
time-varying and spatially dependent.  Experiments that image the
radio sky at multiple frequencies bracketing the plasma cutoff can
therefore simultaneously image the ionosphere and provide spectral
riometry data for space weather studies~\citep{2014GeoRL..41.5370K}.

@@

The back-end readout electronics are housed in a Faraday cage located
50~m away from the antennas. The analog signal chain consists of a
pair of high- and low-pass filters (Mini-Circuits ZFHP-1R2+ and
SLP-150+) that together band-limit the signal to 1.2--140~MHz, and the
filters are followed by a Mini-Circuits ZFL-500+ amplifier.  In
contrast to the two-element pathfinder, the low-pass cutoff is
increased from 81~MHz to 140~MHz to capture downlink signals at
137--138~MHz from the ORBCOMM satellite constellation.  For the final
\albatros\ array, the ORBCOMM signals provide a convenient means for
synchronization across the antenna stations, serving as a backup to
the GPS timing discussed below.  Preliminary lab tests suggest that on
time scales of $\sim30$~s, relative timing between different antenna
stations can be measured to an accuracy of better than a few tenths of
a nanosecond using a single satellite.  The corresponding phase error
at 10~MHz is $\lesssim1^\circ$.  With publicly available orbits and
multiple satellites typically within the field of view, we expect that
ORBCOMM baseband data saved at the same time as the astronomical data
can provide offline synchronization of the \albatros\ stations.
Because the ORBCOMM and science data are recorded at the same time by
the same system, improvements to the timing calibration can be applied
in post-processing.

As with the two-element pathfinder, a SNAP board digitizes each of the
two RF signals at 250~Msamp/s. In the autonomous station
configuration, the SNAP board ADCs are locked to a 10-MHz reference
produced by a Trimble Thunderbolt~E GPS-disciplined clock module.  The
SNAP board FPGA computes two data products: 1) channelized baseband
data for each polarization over tunable frequency windows within the
0--125~MHz operating range, with the options of 1-, 2-, or 4-bit
compression, and 2) auto- and cross-spectra from the two polarizations
over the full 0--125~MHz span, accumulated over few-second intervals.
The reduction in bit depth happens only \textit{after} the SNAP board
has channelized the baseband, and the cross-channel leakage (due to,
e.g., RFI and slopes in RF power as seen by the ADCs) is therefore
unaffected by the low bit depth.  An RPi~3B+ controls the SNAP board
and and receives the auto- and cross-spectra via GPIO connections, and
the spectra are saved to an on-board SD card.  The baseband data are
passed from the SNAP board to the RPi via ethernet and written to
external hard drives.  The introduction of gigabit ethernet with the
RPi~3B+ model has enabled the high data throughput associated with
writing baseband.  As a benchmark, 1-bit baseband recording of two
polarizations over 10~MHz of bandwidth yields an approximate data rate
of 5~MB/s, or 0.4~TB/day.
